Rising from the Ashes II: Did the Birth of the Milky Way's Bar Give Birth to the Birth of the Abundance Bimodality

### 2024-09-07 paper/ms.tex
Update abstract
and change 10.75 to 10.6

\title{Rising from the Ashes II: A Hidden Abundance Bimodality in Illustris TNG}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig/392276_alpha.pdf}
  \caption{\textbf{Bimodality in the abundance plane is linked to distinct epochs in simulation.} The upper panels show \MgFe{} as a function of age for our subhalo in TNG. The colors indicate stellar populations at fixed values of \FeH{}, which are the same as in Figure~\ref{fig:fig1}. A gap in the relation occurs at an age of approximately $10.75\,\Gyr$, which we indicate with a vertical dashed line. The effect of the $\alpha$-enhancement is clear, as it separates the stars that form before and after this gap in ages (star particles which formed before $z=1.5$ are $\alpha$-enhanced, which occurs at an age of $\sim9.5\,\Gyr$). The lower panels show on the left the Milky Way and on the center and right the data from TNG but with $15\%$ age errors and $0.01\,\dex$ errors in \MgFe{}. When the simulations are given these errors, we see that the before and after star particles smear such that the two populations significantly overlap in ages. In the $\alpha$-enhanced SoI, two populations emerge in each bin which overlapped in the fiducial distribution. This feature more closely resembles the Milky Way, which displays such populations where the bimodality is strongest -- $\FeH=-0.5$ (blue) and $-0.25$ (orange).}
  \label{fig:alpha}
\end{figure*}

@@

There is a gap in the ages that occurs at an age of $10.75\,\Gyr$, which we mark with a vertical dashed line. Star particles older than this line have a much clearer gradient in time with \MgFe{} than stars that form after. In the $\FeH=-0.25$ bin, star particles which form directly after this line have a slightly depressed \MgFe{} than stars which form a short time later.

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{fig/392276_SFH_AGN_M200.pdf}
  \caption{\textbf{The evolutionary history of our subhalo of interest.} The upper panel shows the SFH of our subhalo. This SFH is generated using the formation times of all star particles in the subhalo at $z=0$. In this and all subsequent panels we show a vertical dashed line at the same position as in Figure~\ref{fig:alpha} ($\sim3\,\Gyr$), which delineates the separation between the high- and low-$\alpha$ sequences. The gap in ages is naturally associated with a gap in the SFH located at the dashed line. The middle panel shows the BH accretion rate as a fraction of the maximum (Eddington) accretion rate across time. The accretion rate is high early on, but then drops. The gap in ages (dashed line) is coincident with a local peak in the BH accretion rate. In the lower panel we show the mass assembly of the halo, given as the mass within the radius that encloses $200\times$ the mean density of the universe. In this plot mergers are shown as sudden increases in $M_{200}$. Mergers at very early times when the halo is signifcantly less massive than its $z=0$ mass are not shown, but one can see that no clear merger is associated with the gap in stellar ages. Two mergers occur later on at $t\sim10$ and $\sim12\,\Gyr$.}
  \label{fig:history}
\end{figure}

In an effort towards understanding the events in our SoI's history that led to the behavior around the dashed line in Figure~\ref{fig:alpha}, we examine the evolution of some of its key quantities. In the upper panel of Figure~\ref{fig:history}, we show the SFH of our galaxy with the same dashed line as in Figure~\ref{fig:alpha} (at an age of $10.75\,\Gyr$ in that plot is at a time of $\sim3\,\Gyr$ here). This SFH is computed from the formation times of all star particles in the subhalo at $z=0$. This allows us to compute the SFH at fine time resolution. We see two peaks in the SFH at $t\sim2.5$ and $\sim4.5\,\Gyr$ of an amplitude of about $70$ and $60\,\Msunyr$. The dashed line corresponds to a relative drop in the SFR, down to about $10\,\Msunyr$ before quickly recovering.

The middle panel shows the accretion rate of the central black hole as a fraction of the maximum (Eddington) accretion rate. Early in its history ($t<2\,\Gyr$), the subhalo experiences high accretion rates. The accretion rate then steadily declines until $t\sim5\,\Gyr$ at which point it switches to the low accretion radio mode. However, at the dashed line ($\sim3\,\Gyr$) there is a localized peak in the accretion rate, maxing out at about $30\%$ Eddington.

@@

Once the $\alpha$-enhancement post-processing has been done, the SoI that we have studied in this work is consistent with this scenario. The vertical dashed line in Figures~\ref{fig:alpha} and \ref{fig:history} denotes the transition between the high- and low-$\alpha$ sequences. We place it at an age of $10.75\,\Gyr$ or a formation time of $\sim3\,\Gyr$. The upper right panel of Figure~\ref{fig:alpha} shows 2D histograms of \MgFe{} vs star particle age in fixed bins of \FeH{} (each color is a different \FeH{} bin, and an offset is given to \MgFe{} to avoid overlap). Stars which formed before the dashed line had a steep decline of \MgFe{} with time, while the relation in star particles formed after the dashed line is much more flat. At the line, we see a gap in the $\FeH=-0.5$ and $-0.25$ bins.

@@

This bimodality can be traced to an event that occurred at an age of $10.75\,\Gyr$, or formation time of $\sim3\,\Gyr$ ($z\sim2$). Shown as a vertical dashed line in Figure~\ref{fig:alpha}, star particles which form before this form at high-$\alpha$ while stars which form after form at lower $\alpha$. At the dashed line, there is a gap. We argue that the presence of this gap is responsible for the valley in the bimodality (Figure~\ref{fig:fig1}). It is accompanied by a global reduction in the SFR by about a factor of 6 (upper panel of Figure~\ref{fig:history}).

>>>>

\title{Rising from the Ashes II: Did the Birth of the Milky Way's Bar Give Birth to the Birth of the Abundance Bimodality}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig/392276_alpha.pdf}
  \caption{\textbf{Bimodality in the abundance plane is linked to distinct epochs in simulation.} The upper panels show \MgFe{} as a function of age for our subhalo in TNG. The colors indicate stellar populations at fixed values of \FeH{}, which are the same as in Figure~\ref{fig:fig1}. A gap in the relation occurs at an age of approximately $10.6\,\Gyr$, which we indicate with a vertical dashed line. The effect of the $\alpha$-enhancement is clear, as it separates the stars that form before and after this gap in ages (star particles which formed before $z=1.5$ are $\alpha$-enhanced, which occurs at an age of $\sim9.5\,\Gyr$). The lower panels show on the left the Milky Way and on the center and right the data from TNG but with $15\%$ age errors and $0.01\,\dex$ errors in \MgFe{}. When the simulations are given these errors, we see that the before and after star particles smear such that the two populations significantly overlap in ages. In the $\alpha$-enhanced SoI, two populations emerge in each bin which overlapped in the fiducial distribution. This feature more closely resembles the Milky Way, which displays such populations where the bimodality is strongest -- $\FeH=-0.5$ (blue) and $-0.25$ (orange).}
  \label{fig:alpha}
\end{figure*}

@@

There is a gap in the ages that occurs at an age of $10.6\,\Gyr$, which we mark with a vertical dashed line. Star particles older than this line have a much clearer gradient in time with \MgFe{} than stars that form after. In the $\FeH=-0.25$ bin, star particles which form directly after this line have a slightly depressed \MgFe{} than stars which form a short time later.

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{fig/392276_SFH_AGN_M200.pdf}
  \caption{\textbf{The evolutionary history of our subhalo of interest.} The upper panel shows the SFH of our subhalo. This SFH is generated using the formation times of all star particles in the subhalo at $z=0$. In this and all subsequent panels we show a vertical dashed line at the same position as in Figure~\ref{fig:alpha} ($\sim3.2\,\Gyr$), which delineates the separation between the high- and low-$\alpha$ sequences. The gap in ages is naturally associated with a gap in the SFH located at the dashed line. The middle panel shows the BH accretion rate as a fraction of the maximum (Eddington) accretion rate across time. The accretion rate is high early on, but then drops. The gap in ages (dashed line) is coincident with a local peak in the BH accretion rate. In the lower panel we show the mass assembly of the halo, given as the mass within the radius that encloses $200\times$ the mean density of the universe. In this plot mergers are shown as sudden increases in $M_{200}$. Mergers at very early times when the halo is signifcantly less massive than its $z=0$ mass are not shown, but one can see that no clear merger is associated with the gap in stellar ages. Two mergers occur later on at $t\sim10$ and $\sim12\,\Gyr$.}
  \label{fig:history}
\end{figure}

In an effort towards understanding the events in our SoI's history that led to the behavior around the dashed line in Figure~\ref{fig:alpha}, we examine the evolution of some of its key quantities. In the upper panel of Figure~\ref{fig:history}, we show the SFH of our galaxy with the same dashed line as in Figure~\ref{fig:alpha} (at an age of $10.6\,\Gyr$ in that plot is at a time of $\sim3.2\,\Gyr$ here). This SFH is computed from the formation times of all star particles in the subhalo at $z=0$. This allows us to compute the SFH at fine time resolution. We see two peaks in the SFH at $t\sim2.5$ and $\sim4.5\,\Gyr$ of an amplitude of about $70$ and $60\,\Msunyr$. The dashed line corresponds to a relative drop in the SFR, down to about $10\,\Msunyr$ before quickly recovering.

The middle panel shows the accretion rate of the central black hole as a fraction of the maximum (Eddington) accretion rate. Early in its history ($t<2\,\Gyr$), the subhalo experiences high accretion rates. The accretion rate then steadily declines until $t\sim5\,\Gyr$ at which point it switches to the low accretion radio mode. However, at the dashed line ($\sim3.2\,\Gyr$) there is a localized peak in the accretion rate, maxing out at about $30\%$ Eddington.

@@

Once the $\alpha$-enhancement post-processing has been done, the SoI that we have studied in this work is consistent with this scenario. The vertical dashed line in Figures~\ref{fig:alpha} and \ref{fig:history} denotes the transition between the high- and low-$\alpha$ sequences. We place it at an age of $10.6\,\Gyr$ or a formation time of $\sim3.2\,\Gyr$. The upper right panel of Figure~\ref{fig:alpha} shows 2D histograms of \MgFe{} vs star particle age in fixed bins of \FeH{} (each color is a different \FeH{} bin, and an offset is given to \MgFe{} to avoid overlap). Stars which formed before the dashed line had a steep decline of \MgFe{} with time, while the relation in star particles formed after the dashed line is much more flat. At the line, we see a gap in the $\FeH=-0.5$ and $-0.25$ bins.

@@

This bimodality can be traced to an event that occurred at an age of $10.6\,\Gyr$, or formation time of $\sim3.2\,\Gyr$ ($z\sim2$). Shown as a vertical dashed line in Figure~\ref{fig:alpha}, star particles which form before this form at high-$\alpha$ while stars which form after form at lower $\alpha$. At the dashed line, there is a gap. We argue that the presence of this gap is responsible for the valley in the bimodality (Figure~\ref{fig:fig1}). It is accompanied by a global reduction in the SFR by about a factor of 6 (upper panel of Figure~\ref{fig:history}).
